Mergeable Types

### 2017-02-25 paper/mergeable-types.tex
edits

\begin{figure}
\centering
\includegraphics[scale=0.4]{Figures/pushouts}

\caption{Commutative diagram representing the merge operation.}
\label{fig:pushouts}
\end{figure}

We now formally define what we mean by a mergeable type. The
definition serves as a guide to write merge functions for functional
data types, and elevate them to the level of replicated data types.

A functional data type library is as a tuple consisting of a type
($t$), and a (possibly infinite) collection of functions ($f$, $g$,
etc.) of type $t \rightarrow t$. We call the functions \emph{primitive
morphisms}\footnote{We borrow terminology from category theory for
convenience}. A morphism ($P$, $Q$, etc.) is either a primitive
morphism, or an associative composition of morphisms ($P \circ Q \circ
R$). An object $B:t$ is called a \emph{semantic successor} of $A:t$
(conversely, $A$ is called a \emph{semantic ancestor} of $B$) if and
only if there exists a morphism $P$ such that $P(A) = B$.

The aim of a three-way merge function over a type $t$ is to merge a
pair of semantic successors, $B:t$ and $C:t$, of an object $A:t$, into
another object $D:t$ such that the relationship between the semantic
successors and $D$ satisfies certain conditions. These conditions can
be understood by observing the \rulelabel{E-Pull-Wait} rule of of
Fig.~\ref{fig:opsem}, which applies the \C{merge} function to a pair
of concurrent versions ($v_1$ and $v_2$) and their least common
ancestor ($v$). Thus, the only relationship that exists between $v$
and $v_1$, and $v$ and $v_2$ is the syntactic ancestor relationship
that follows from the branching structure. The merge function, as
described above, however assumes that the concurrent versions are
semantic successors of their LCA. It is therefore essential to
maintain coherence between the syntactic and semantic ancestor
relations.

The ways in which syntactic ancestor relationships are created among
versions is captured in Fig.~\ref{fig:syntactic-ancestors}. Whenever an
object is pushed onto the branch, an ancestor relationship is created
between the previous version $v_1$, and the newly created version
$v_2$ (Fig.~\ref{fig:syntactic-ancestor-1}). However, since $v_2$ is
pushed by the thread after reading $v_1$ into the heap, it is
reasonable to assume that $v_2$ is a result of applying a morphism $P$
to $v_1$ (i.e., $\exists P.~v_2 = P(v_1)$).  Hence, $v_1$ is a
semantic ancestor of $v_2$.  Fig.~\ref{fig:syntactic-ancestor-2}
captures another way of establishing ancestor relationship, namely by
merging branches. Version $v_{21}$ on branch $b_2$ is merged into
$v_{11}$ on $b_1$ to create $v_{12}$ on $b_1$. Versions $v_{11}$ and
$v_{21}$ are now syntactic ancestors of $v_{12}$, but for them to be
semantic ancestors, merge function has to enforce the relationship
explicitly. In other words, the result of merging of a pair of
semantic successors, $B$ and $C$, of an $A$, has to be an object $D$
that is a semantic successor of both $B$ and $C$. We now formalize
this intuition to define a mergeable type.
\begin{definition} [\bfseries Mergeable Type]
\label{def:mergeable-type}
A type $t$ is said to be mergeable, if and only if there exists a
function $M$ of type $t \times t \times t \rightarrow (t \rightarrow
t)\times(t \rightarrow t)$ that satisfies the following property:
$\forall (A, B, C\,:\,t).~ (\exists (P,Q\,:\, t \rightarrow t).~ B =
P(A) \conj C = Q(A)) \Rightarrow (\exists (P',Q': t\rightarrow
t).~M(A,B,C) = (P',Q') \conj  Q'(B) = P'(C))$
\end{definition}
Intuitively, a type $t$ is a mergeable type if, whenever there exist
two morphisms $P$ and $Q$ that map $A:t$ to $B:t$ and $C:t$, there
must also exist morphisms $P'$ and $Q'$, the map both $B$ and $C$ to
$D:t$. This intuition is expressed visusally in
Fig.~\ref{fig:pushouts}.

For a mergeable type $t$, we have the guarantee that a sequence of
values read by a thread from its branch ``makes sense'' as per the
data type semantics:

@@

Def.~\ref{def:mergeable-type} specifies the merge function ($M$)
slightly differently; instead of returning a merged value like the
three-way merge function (\C{merge}), $M$ returns a pair of morphisms.
This specification is useful to preempt certain non-sensical merge
functions, and thus enforce the branch-local consistency property
(Theorem~\ref{thm:branch-consistency}). For instance, consider the
counter data type from Sec.~\ref{sec:intro}. A three-way merge
function can be defined to return a constant regardless of its
arguments:
\begin{ocaml}
  let merge lca a b = 0
\end{ocaml} Such a merge function allows the client (a thread) to
witness the violation of the branch-local consistency property. For
example, the client may \C{sync} the counter, whose current local
value is $10$, and obtain $0$ as a result, thus witnessing a violation
of counter's monotonicity property. Fortunately,
Def.~\ref{def:mergeable-type} disallows this semantics. It is
impossible to define the function $M$ for the counter data type that
violates the monotonicity invariant, because no counter morphism
violates the invariant. In general, the specification of $M$
guarantees that the merge operation for a type $t$ preserves any
invariants (e.g., balancedness of a tree, sortedness of a list,
non-negativeness of an integer etc) that are preserved by all of $t$'s
morphisms. On the hand, the commutativity of the diagram in
Fig.~\ref{fig:pushouts} may not be enforced statically. It can however
be verified at run-time by checking that the morphisms returned by $M$
map the respective concurrent versions to the same value. The
operational semantics (Fig.~\ref{fig:opsem}) can be extended to this
effect.

@@

Merge functions for simple data types are easy to write. For instance,
the merge logic for counter is simple enough that it doesn't require
explicit efforts to identify $P$ and $Q$ that led to the
concurrent counter versions, and transform them into $P'$ and $Q'$
such that the diagram in Fig.~\ref{fig:pushouts} commutes. For more
sophisticated data types, this process is more involved, and requires
the programmer to perform non-trivial reasoning. We now present an
alternative definition of a mergeable type that is stronger than
Def.~\ref{def:mergeable-type}, but also serves as a recipe to build
non-trivial merge functions.

@@

\paragraph{Strings} Strings are mergeable. Given a string ($s$) and
its semantic successor ($s'$), an edit script ($P^*$ or $Q^*$) is the
shortest sequence of \C{insert}, \C{delete}, and \C{substitute}
operations (string's primitive morphisms), whose composition takes $s$
to $s'$. The length of the sequence is in fact the Levenshtein
distance~\cite{levenstein} between $s$ and $s'$, which can be computed
in polynomial time via Wagner-Fischer algorithm~\cite{wagner-fischer}.
A slight modification of Wagner-Fischer also lets us compute the edit
script. Fig.~\ref{fig:mergeable-string-eg} shows the edit script ...

>>>>

We now formally define what we mean by a mergeable type. The
definition serves as a guide to write merge functions for functional
data types, elevating them to the level of replicated data types.

A functional data type library is a tuple consisting of a type ($t$),
and a collection of functions ($f$, $g$, etc.) of type $t \rightarrow
t$ that are \emph{primitive morphisms}. A morphism ($P$, $Q$, etc.) is
either a primitive morphism, or an associative composition of
morphisms ($P \circ Q \circ R$). An object $B:t$ is called a
\emph{semantic successor} of $A:t$ (conversely, $A$ is called a
\emph{semantic ancestor} of $B$) if and only if there exists a
morphism $P$ such that $P(A) = B$.

The aim of a three-way merge function over a type $t$ is to merge a
pair of semantic successors, $B:t$ and $C:t$, of an object $A:t$, into
another object $D:t$ such that the relationship between the semantic
successors and $D$ satisfies certain conditions. These conditions can
be understood by observing the \rulelabel{E-Pull-Wait} rule of
Fig.~\ref{fig:opsem}, which applies the \C{merge} function to a pair
of concurrent versions ($v_1$ and $v_2$) and their least common
ancestor ($v$). Thus, the only relationship that exists between $v$
and $v_1$, and $v$ and $v_2$ is the syntactic ancestor relationship
that follows from the computation's branching structure. The merge
function, as described above, however assumes that concurrent versions
are semantic successors of their LCA.  It is therefore essential to
maintain coherence between the syntactic and semantic ancestor
relations.

The ways in which syntactic ancestor relationships are created among
versions is captured in Fig.~\ref{fig:syntactic-ancestors}. Whenever an
object is pushed onto the branch, an ancestor relationship is created
between the previous version $v_1$, and the newly created version
$v_2$ (Fig.~\ref{fig:syntactic-ancestor-1}). However, since $v_2$ is
pushed by the thread after reading $v_1$ into the heap, it is
reasonable to assume that $v_2$ is a result of applying a morphism $P$
to $v_1$ (i.e., $\exists P.~v_2 = P(v_1)$).  Hence, $v_1$ is a
semantic ancestor of $v_2$.  Fig.~\ref{fig:syntactic-ancestor-2}
captures another way of establishing an ancestor relationship, namely by
merging branches. Version $v_{21}$ on branch $b_2$ is merged into
$v_{11}$ on $b_1$ to create $v_{12}$ on $b_1$. Versions $v_{11}$ and
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.35]{Figures/pushouts}
\caption{Commutative diagram representing the merge operation.}
\label{fig:pushouts}
\end{wrapfigure}
$v_{21}$ are now syntactic ancestors of $v_{12}$, but for them to be
semantic ancestors, the merge function has to enforce the relationship
explicitly. In other words, the result of merging  a pair of
semantic successors, $B$ and $C$, of an $A$, has to be an object $D$
that is a semantic successor of both $B$ and $C$. We now formalize
this intuition to define a mergeable type.
\begin{definition} [\bfseries Mergeable Type]
\label{def:mergeable-type}
A type $t$ is said to be mergeable, if and only if there exists a
function $M$ of type $t \times t \times t \rightarrow (t \rightarrow
t)\times(t \rightarrow t)$ that satisfies the following property:
$\forall (A, B, C\,:\,t).~ (\exists (P,Q\,:\, t \rightarrow t).~ B =
P(A) \conj C = Q(A)) \Rightarrow (\exists (P',Q': t\rightarrow
t).~M(A,B,C) = (P',Q') \conj  Q'(B) = P'(C))$
\end{definition}
Intuitively, a type $t$ is a mergeable type if, whenever there exist
two morphisms $P$ and $Q$ that map $A:t$ to $B:t$ and $C:t$, there
also exists morphisms $P'$ and $Q'$, that map both $B$ and $C$ to
$D:t$. This intuition is expressed visually in
Fig.~\ref{fig:pushouts}.  For a mergeable type $t$, we have the
guarantee that a sequence of values read by a thread from its branch
is sensible, as per the data type semantics:

@@

Rather than specifying merge ($M$) in terms of an LCA,
Definition~\ref{def:mergeable-type} specifies $M$ in terms of a pair
of morphisms.  This specification is useful to avoid the expression of certain
nonsensical merge functions, and thus enforce a useful branch-local
consistency property (Theorem~\ref{thm:branch-consistency}). For
instance, consider the counter data type from Sec.~\ref{sec:intro}. A
three-way merge function can be defined to return a constant
regardless of its arguments:
\begin{ocaml} let merge lca a b = 0 \end{ocaml}
Such a merge function allows the client to
witness the violation of the branch-local consistency property. For
example, the client may \C{sync} the counter, whose current local
value is $10$, and obtain $0$ as a result, thus witnessing a violation
of counter's monotonicity property. Fortunately,
Def.~\ref{def:mergeable-type} disallows this semantics. It is
impossible to define the function $M$ for the counter data type that
violates the monotonicity invariant, because no counter morphism
violates the invariant. In general, the specification of $M$
guarantees that the merge operation for a type $t$ preserves any
invariants (e.g., balancedness of a tree, sortedness of a list,
non-negativeness of an integer etc) that are preserved by all of $t$'s
morphisms.  The commutativity of the diagram in
Fig.~\ref{fig:pushouts} need not be enforced statically. It can
equally be verified at run-time by checking that the morphisms returned by $M$
map respective concurrent versions to the same value. The
operational semantics (Fig.~\ref{fig:opsem}) can be extended to this
effect.

@@

Merge functions for simple data types are easy to write. For instance,
the merge logic for counter is simple enough that it does not require
great effort to identify the $P$ and $Q$ that leda to the
concurrent counter versions, and transform them into $P'$ and $Q'$
such that the diagram in Fig.~\ref{fig:pushouts} commutes. For more
sophisticated data types, this process is more involved, and may require
the programmer to perform non-trivial reasoning. We now present an
alternative definition of a mergeable type that is stronger than
Def.~\ref{def:mergeable-type}, but also serves as a recipe to build
non-trivial merge functions.

@@

\paragraph{Strings} Strings are mergeable. Given a string ($s$) and
its semantic successor ($s'$), an edit script ($P^*$ or $Q^*$) is the
shortest sequence of \C{insert}, \C{delete}, and \C{substitute}
operations (a string's primitive morphisms), whose composition takes $s$
to $s'$. The length of the sequence is in fact the Levenshtein
distance~\cite{levenstein} between $s$ and $s'$, which can be computed
in polynomial time~\cite{wagner-fischer}.
A slight modification of Wagner-Fischer also lets us compute the edit
script. Fig.~\ref{fig:mergeable-string-eg} shows the edit script ...

### 2017-02-25 paper/meta.tex
edits

Ancestor relation is therefore a partial order (reflexive, transitive,
anti-symmetric) with a greatest lower bound (the initial version).
Thus, for any two versions in a legal history, there exist at least
one common ancestor. Ancestor relationships among the common ancestors
let us define the notion of a least common ancestor (LCA):

@@

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/merge-needs-lca}

\caption{This example of a grow-only counter illustrate why \C{merge}
needs a least common ancestor, and not just a common ancestor. Both 0
and 2 are common ancestors of 4 and 7, while 2 is their least common
ancestor (since $0 \preceq 2$). The result (v) of merging 4 and 7 is
11 (incorrect) if 0 is used as the common ancestor for merge, and 9
(correct, because 2+2+3+2 = 9) if 2 is used. }
\label{fig:merge-needs-lca}
\end{figure}

@@

When merging two concurrent versions $v_1$ and $v_2$, the common
ancestor argument for \C{merge} must the LCA of $v_1$ and $v_2$,
without which \C{merge} may yeild unexpected results. This is
demonstrated for the grow-only counter in
Fig.~\ref{fig:merge-needs-lca}, where incorrect count is obtained if a
common ancestor that is not an LCA is used to merge 4 and 7. While in
this example there is a unique LCA for 4 and 7, in general this may
not be the case. With unrestrained branching and merging, there is no
bound on the number of LCAs a pair of versions can have.
For example, in
Fig.~\ref{fig:criss-cross-lcas}, the merge of 0 with 3 is preceded by
two ``criss-cross'' merges between their respective branches resulting
in there being two LCAs (5 and 4) for 0 and 3. Multiple LCAs can occur
even without criss-cross merges, as demonstrated by
Fig.~\ref{fig:external-lcas}. Concurrent versions with multiple LCAs
do not lend themselves to three-way merging. This problem also arises
in the context of source control systems, which employ ad hoc
mechanisms to pave way for three-way merging.  GitHub~\cite{github},
for instance, recursively merges LCAs to compute a virtual ancestor,
which then serves as the LCA for merging the concurrent versions. This
method is demonstrated for the example in
Fig.~\ref{fig:criss-cross-lcas}, where LCAs 5 and 4 of 0 and 3 are
merged (with LCA of 10) to generate -1 as the virtual LCA to merge 5
and 4. A major downside with this method is that it makes no
guarantees of the relationship between the virtual ancestor and the
concurrent versions; the former may not even be a legal ancestor of
the latter as per the semantics of the data type. For instance, the
integer type in Fig.~\ref{fig:criss-cross-lcas} may in fact represent
a bank account, which disallows any activity on the account if balance
is ever known to be less than zero. Thus, from the perspective of a
bank account, it doesn't make sense how concurrent versions 3 and 0
emerged from -1, since the only transition allowed by the semantics
from -1 is to itself. Clearly, ad hoc mechanisms, which work for the
text, may not work for more sophisticated data types.

Fortunately, unlike the source control systems where branching
structure is entirely dictated by the user, \name abstracts away the
branching structure from the programmer, hence retains the ability to
manifest it in a way that it deems fit. In particular, \name solves
the problem of multiple LCAs either by constraining the branching
stucture to preempt multiple LCAs, or by extending it such that
merging branches have a new unique LCA. The preemption is done for
criss-cross merges by the operational semantics, which only allows
merging versions that are latest on their respective branches. For
instance let $v_{11}$ and $v_{21}$ be the latest versions on branches
$b_1$ and $b_2$, respectively. Merging $b_2$ into $b_1$ entails
merging $v_{21}$ into $v_{11}$ to generate version $v_{12}$ on $b_1$.
Now, merging $b_1$ into $b_2$ involves merging $v_{12}$, the latest
version on $b_1$, into $v_{21}$, but not $v_{11}$ into $v_{21}$, thus
preempting a criss-cross branching structure. 
% In other words, a criss-cross branching structure is prevented due
% to the order among merges between conflicting branches introduced as
% a result of \rulelabel{E-Pull-Wait} being an atomic step.

Next, for the branching structures where a pair of branches (i.e.,
their respective latest versions) have multiple externally-located
LCAs, the problem can be fixed by suitably extending the branching
structure such that the branches have a new single LCA. For instance,
consider the example in Fig.~\ref{fig:external-lcas}, where (latest
versions on) branches $b_1$ and $b_4$ have two externally-located LCAs
($v_{22}$ and $v_{32}$), hence cannot be merged. Mergeability can
however be reclaimed by extending the branching structure as shown in
Fig.~\ref{fig:legal-extension-1}. The extension is done by first merging
$b_3$ into $b_2$ to create version $v_{23}$ on $b_2$, then merging
$b_2$ into $b_1$ to create version $v_{14}$ on $b_1$, and finally
merging $b_2$ into $b_4$ to create version $v_{45}$ on $b_4$. Observe
that $b_1$ and $b_4$ now have a single LCA ($v_{23}$) hence are now
mergeable. Mergeability could also have been reclaimed by adopting an
alternative extension to the branching structure as shown in
Fig.~\ref{fig:legal-extension-2}.

>>>>

The ancestor relation is therefore a partial order with a greatest
lower bound (the initial version).  Thus, for any two versions in a
legal history, there exists at least one common ancestor.  The ancestor
relationship among all common ancestors lets us define the notion of a
least common ancestor (LCA):

@@

When merging two concurrent versions $v_1$ and $v_2$, the common
ancestor argument for \C{merge} must be the LCA of $v_1$ and $v_2$,
without which \C{merge} may yield unexpected results. This is
demonstrated for the grow-only counter in
Fig.~\ref{fig:merge-needs-lca}, where an incorrect count is obtained if a
common ancestor that is not an LCA is used to merge 4 and 7. While in
this example there is a unique LCA for 4 and 7, in general this may
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.6]{Figures/merge-needs-lca}
\caption{This example of a grow-only counter illustrate why \C{merge}
needs a least common ancestor, and not just a common ancestor. Both 0
and 2 are common ancestors of 4 and 7, while 2 is their least common
ancestor (since $0 \preceq 2$). The result (v) of merging 4 and 7 is
11 (incorrect) if 0 is used as the common ancestor for merge, and 9
(correct, because 2+2+3+2 = 9) if 2 is used. }
\label{fig:merge-needs-lca}
\end{wrapfigure}
not be the case. With unrestrained branching and merging, there is no
bound on the number of LCAs a pair of versions can have.  For example,
in Fig.~\ref{fig:criss-cross-lcas}, the merge of 0 with 3 is preceded
by two ``criss-cross'' merges between their respective branches
resulting in there being two LCAs (5 and 4) for 0 and 3. Multiple LCAs
can occur even without criss-cross merges, as demonstrated by
Fig.~\ref{fig:external-lcas}. Concurrent versions with multiple LCAs
do not lend themselves to three-way merging. This problem also arises
in the context of source control systems, which employ \emph{ad hoc}
mechanisms to pave the way for three-way merging.
GitHub~\cite{github}, for instance, recursively merges LCAs to compute
a virtual ancestor, which then serves as the LCA for merging
concurrent versions. This method is demonstrated in the branching
structure for a mergeable, replicated counter as shown in
Fig.~\ref{fig:criss-cross-lcas}, where LCAs 5 and 4 of 0 and 3 are
merged (with their LCA being 10) to generate -1 as the virtual LCA to
merge 5 and 4. A major downside with this approach is that it makes no
guarantees on the relationship between the virtual ancestor and its
concurrent versions; the former may not even be a legal ancestor of
the latter as per the semantics of the data type. For instance,
suppose the integer type in Fig.~\ref{fig:criss-cross-lcas} represents
a bank account balance, which is expected to disallows any activity on
the account if the balance is ever known to be less than zero.  From
the perspective of the library designer and its clients, there is no
meaningful scenario in which versions 3 and 0 can emerge from -1,
since the only transition allowed by the semantics from -1 is to
itself.  Clearly, \emph{ad hoc} mechanisms like this are error-prone
and difficult to apply in general.

@@

Fortunately, unlike source control systems where branching structure
is entirely dictated by the user, \name abstracts away branching
structure from the programmer, and hence retains the ability to
manifest it in a way that it deems fit. In particular, \name solves
the problem of multiple LCAs either by constraining the branching
stucture to prevent their existence, or by merging branches to ensure
that they have a new unique LCA.  Preention is done for
criss-cross merges by the operational semantics, which only allows
merging versions that are the latest on their respective branches. For
instance, suppose $v_{11}$ and $v_{21}$ are the latest versions on branches
$b_1$ and $b_2$, respectively. Merging $b_2$ into $b_1$ entails
merging $v_{21}$ into $v_{11}$ to generate version $v_{12}$ on $b_1$.
Now, merging $b_1$ into $b_2$ involves merging $v_{12}$, the latest
version on $b_1$, into $v_{21}$, but not $v_{11}$ into $v_{21}$, thus
preempting a criss-cross branching structure.
% In other words, a criss-cross branching structure is prevented due
% to the order among merges between conflicting branches introduced as
% a result of \rulelabel{E-Pull-Wait} being an atomic step.

Next, for branching structures where a pair of branches (i.e., their
respective latest versions) have multiple externally-located LCAs, we
suitably extend the branching structure to generate a new single LCA.
 For instance, consider
the example in Fig.~\ref{fig:external-lcas}, where (latest versions
on) branches $b_1$ and $b_4$ have two externally-located LCAs
($v_{22}$ and $v_{32}$), hence cannot be merged. Mergeability can
however be reclaimed by extending the branching structure as shown in
Fig.~\ref{fig:legal-extension-1}. The extension is done by first
merging $b_3$ into $b_2$ to create version $v_{23}$ on $b_2$, then
merging $b_2$ into $b_1$ to create version $v_{14}$ on $b_1$, and
finally merging $b_2$ into $b_4$ to create version $v_{45}$ on
$b_4$. Observe that $b_1$ and $b_4$ now have a single LCA ($v_{23}$)
hence are now mergeable. Mergeability could also have been reclaimed
by adopting an alternative extension to the branching structure as
shown in Fig.~\ref{fig:legal-extension-2}.
